Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-04-25 Paper/Paper.tex

Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation. Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching. Whether physicians are retirement age or not, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them. In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group. I observe similar trends in whether physicians change zip codes, although pre-trends may be driving this result. Finally, I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR. The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.

@@

The first dependent variable I consider is the decision for a physician to retire, or more broadly to stop practicing in a clinical setting altogether. Retirement is constructed as follows: 
$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$. Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future \footnote{Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results. In Appendix \textcolor{red}{()}, I limit the sample years to 2009-2015 to ensure sufficient future year of zero claims.}.

Next, I consider whether physicians exhibit behavior that suggests they have shifted towards other work settings. I measure this in the data in three ways. First, I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data. Second, I construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office-based setting. This binary measure of practice setting offers less variation but captures a more discrete changes in which physicians enter or leave the office-based setting entirely. Third, based on the physician's practice zip code, I form an indicator variable equal to one if the physician's list of zip codes changes from one year to the next. While no single variable directly measures physician movement across hospitals, these three outcomes collectively provide insight into whether physicians desire to shift away from EHRs.

@@

I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}. The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean. This is positive, but somewhat inconsequential effect, both economically and compared to the effect of EHRs on retirement discussed above. Splitting the sample by age reveals that this result is driven by the younger sample of physicians. There is no evidence that older physicians are shifting towards offices. This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more "stuck" to their particular work environment.

@@

The effects of EHR exposure on zip code switching over time are shown in Figure \ref{fig:zip}. The graph behaves similarly to the probability of working in an office in that the probability of switching zip codes increases and the result is driven solely by the younger demographic. However, this result is meaningfully larger; a physician is .025 ppts more likely to switch zip codes in the same year as EHR exposure and the year after, a 31.25\% increase relative to the mean. This result suggests that EHRs are imposing a cost high enough for physicians to exhibit significant switching behaviors, whether that is towards offices or other hospitals.

@@

An important, and heavily debated, aspect of the implementation of EHRs is whether or not they allow users to be more productive. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden. Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments. Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all. In this section, I address EHRs' effect on productivity by estimating its effect on a physician's number of patients seen.

@@

This variable is also driven by demand for health care. A major change in the U.S. health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook. If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, the ACA may be driving the results. The ACA could potentially affect patient counts in opposite ways: if Medicare patients are crowded out by an increased demand in Medicaid patients, or if the small changes to Medicare led to an increase in utilization from the Medicare population. A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}). Further, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}. Each group shows the same pattern of a small increase in patient count in the first 0-2 years after implementation.

@@

Further, EHRs also have the potential to change billing activity differentially than patient count changes. If, for example, EHRs reduce the administrative burden of filing extra claims, then physicians might increase the things they bill for whether the amount of patients has changed or not. On average for the entire sample, claim counts are roughly 5.5 times the number of patients seen in a year. Shown in Figure \ref{fig:claim}, young physicians increase claim count by 37 in the year of exposure and 96 in the year after exposure, 3-4 times the increase in patient count. Thus, billing activity increases proportionally less than average. EHRs do not seem to be facilitating more billable activity.

However, for both patient and claim count, the results include physicians who shift place of work, which may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR. Thus, I limit the sample to only physicians who work stay with the same one hospital through the entire sample. These physicians do not retire or see patients in another hospital. Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the period. I present the results for this limited sample in Figure \ref{fig:limitedsample}.

@@

In all outcomes, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure. All specifications point to the effect of EHR exposure to become null in the 3-4 years after exposure. This is likely due to the composition of the hospitals adopting at different times. Consider time $t+4$. The only hospitals to analyze the effect in this period are those who adopted in 2010 compared to hospitals who adopted in 2015. The group of hospitals who adopted in 2015 may be compositionally different than early adopters if there is a characteristic reason for late adoption such as demographic of patients. For this reason, I focus discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.

@@

One concern with the presented analysis is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals. One such occurrence would be the existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf, which is a way for physicians to continue working in a hospital with an EHR but not actually use the EHR. Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals. they have the following titles: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator. I will refer to any employee in these categories as a data assistant. In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.

First, using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals. Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above. The total number of these registered employees is 875, and the first data assistants ever registered were in 2005. From 2005 to 2013, 15-60 more employees were registered each year. The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014. If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs. My results are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.

@@

@@

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Claim Count}
    \includegraphics[scale=.4]{Objects/claim_plot.pdf}
    \label{fig:claim}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Productivity, Limited Sample}
    \includegraphics[scale=.4]{Objects/limitedsample_plot.pdf}
    \label{fig:limitedsample}
\end{figure}

@@

I use the general survey question about each hospital's EHR use to define my key treatment variable. The survey asks the extent to which the hospital utilizes an EHR and allows for three answers: not at all, partially utilize an EHR, and fully utilize an EHR. I create a binary variable equal to one if the answer is fully utilize, zero otherwise. The data will be aggregated up to the physician level due to physician level outcome variables, so I first transform the pair-level EHR variable into a treatment variable at the physician level. I create a variable capturing the first year that at least one of a physician's connected hospitals uses an EHR. For the endogeneity discussion in Section \ref{sec:endog}, I create a similar variable but only count a physician as exposed if a hospital categorized as "low-integration" adopts an EHR. The level of vertical integration between hospital and physician is defined as in Madison (\citeyear{madison2004hospital}) using the organization type of the hospital. Hospitals classified as IPA or PHO are low-integration, and any other type is high0integration. I also create a variable that captures the fraction of hospitals a physician works with that use an EHR, which varies on the dimension of EHR use and on number of hospitals worked with. Finally, the last variable I create before aggregating to the physician level is an indicator for whether the physician works with the same set of hospitals through the entire sample. I create this variable by comparing a physician's number of hospitals in each year to the maximum number of hospitals they ever work with. If there is any year in which the a physician's number of hospitals is less than the maximum number of hospitals they ever work with, the indicator for never having a new NPI is set to 0. I use this variable to limit the sample in Section \ref{sec:patientcount}.

@@

The main specification of the paper defines a binary treatment variable for whether a physician is exposed to an EHR in any hospital they work in. A concern that one may have is that physicians choose their individual EHR exposure, making this an endogenous treatment variable. The results do not indicate that this is happening on average, since we would expect physicians not to respond on the margin of retirement or switching zip codes if they are choosing to be exposed to EHRs. Especially in larger hospitals, this is unlikely to occur as an individual physician will not have much input to high level decisions. However, as an attempt to provide more evidence of labor market changes as a result of EHR exposure, I provide results for a specification with an alternate binary treatment variable. This variable is whether a physician has been exposed to an EHR in a hospital considered to have low vertical integration with physicians.

@@

>>>>

Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation. Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching or avoiding Medicare patients. Whether physicians are retirement age or not, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them. In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group. Finally, I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR. The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.

@@

The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I loosely call retirement. This variable is constructed as follows: 
$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$. Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future \footnote{Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results. In Appendix \textcolor{red}{()}, I limit the sample years to 2009-2015 to ensure sufficient future year of zero claims.}.

Next, I consider whether physicians exhibit behavior that suggests they have shifted setting or location of work. I measure this in the data in three ways. First, I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data. Second, I construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office-based setting. This binary measure of practice setting offers less variation but captures a more discrete changes in which physicians enter or leave the office-based setting entirely. Third, based on the physician's practice zip code, I form an indicator variable equal to one if the physician's list of zip codes changes from one year to the next. While no single variable directly measures physician movement across hospitals, these three outcomes collectively provide insight into whether physicians desire to shift away from EHRs.

@@

I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}. The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean. This is positive, but somewhat inconsequential effect, both economically and compared to the effect of EHRs on retirement discussed above. Splitting the sample by age reveals that this result is driven by the younger sample of physicians. There is no evidence that older physicians are shifting towards offices. This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more stuck to their particular work environment.

@@

The effects of EHR exposure on zip code switching over time are shown in Figure \ref{fig:zip}. While there is a statistically positive effect in the first 2 years of exposure, there is an upward sloping pre-trend that is probably driving this. Thus, I make no claim that EHR exposure is leading physicians to switch zip codes.

A limitation in this analysis is that physicians may shift work behavior due to the organization of the hospital system changing. For example, a hospital may acquire a smaller practice and send physicians to work there for a period of time. If this acquisition is correlated with incentives to adopt an EHR, then the effect I find is biased upward. Since I do not find strong evidence for switching behavior, I leave this question to future research.

@@

An important, and heavily debated, aspect of the implementation of EHRs is whether or not they allow users to be more productive. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden. Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments. Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all. In this section, I speak to efficiency by estimating the effect of EHR implementation on patient count.

@@

Patient count is also driven by demand for health care. A major change in the U.S. health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook. If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, the ACA may be driving the results. The ACA has potentially opposing effects on Medicare patients: if Medicare patients are crowded out by an increased demand in Medicaid patients, or if the small changes to Medicare led to an increase in utilization from the Medicare population. A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results. Further, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}. Each group shows the same pattern of a small increase in patient count in the first 0-2 years after implementation.

@@

EHRs may affect physician incentives in such a way that billing activity changes differently from patient count. If, for example, EHRs reduce the administrative burden of filing extra claims, then physicians might increase the things they bill for whether the amount of patients has changed or not. On average for the entire sample, claim counts are roughly 5.5 times the number of patients seen in a year. Shown in Figure \ref{fig:claim}, young physicians increase claim count by 37 in the year of exposure and 96 in the year after exposure, 3-4 times the increase in patient count. Thus, billing activity increases proportionally less than average. EHRs do not seem to be facilitating more billable activity.

However, for both patient and claim count, the results include physicians who shift place of work, which may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR. Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample. These physicians do not retire or see patients in another hospital. Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the period. I present the results for this limited sample in Figure \ref{fig:limitedsample}. These results are similar to those presented for the full sample, suggesting that physicians become more productive in terms of patient count in the first two years after implementation.

@@

In the observed results above, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure. Almost all specifications point to the effect of EHR exposure to become null in the 3-4 years after exposure. One reason for this may be the composition of the hospitals which adopt at different times. Consider event time $t+4$. The physician included in the estimation for this coefficient are those exposed in 2010 compared with those who did not adopt until 2015, implying that the underlying hospitals implemented EHRs in either 2010 or 2015. It is likely that the group of hospitals who adopted in 2015 are compositionally different than early adopters, especially if poorer hospitals adopted later. For this reason, I have focused discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.

@@

One concern with the presented analysis is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals, such as the existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf, which is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology. Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals. they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator. I will refer to any employee in these categories as a data assistant. In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.

First, using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals. Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above. The total number of these registered employees is 875, and the first data assistants ever registered were in 2005. From 2005 to 2013, 15-60 more employees were registered each year. The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014. If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs. The results of my main specification are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.

@@

Another way physicians may avoid EHR exposure is to exhibit decision making power to either delay or encourage EHR implementation in hospitals. This would make the binary treatment variable endogenous, biasing the results of the main specification, as it includes physicians who are selecting into their own exposure. After the sample period ends, the government began penalizing hospitals for not using health information technology meaningfully, so implementation eventually occurs across the board. However, during the sample, physicians may impact the timing or occurrence of implementation. I address this question by limiting the sample of hospitals in a meaningful way, only including those hospitals which are considered to have low vertical integration with physicians. That is, physicians are employees and less likely to be decision makers. I then redefine exposure as when a physician is exposed to an EHR in low integration settings. I present details of how I define integration and results for this specification in Appendix \ref{sec:endog}. In general, group time estimates are similar to those in the main specification, with the exception of the probability of working in an office setting, which may indicate some correlation between EHR adoption in high integration settings and acquisition of physician offices.

@@

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Claim Count, Limited Sample}
    \includegraphics[scale=.4]{Objects/claim_plot.pdf}
    \label{fig:claim}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Patient Count, Limited Sample}
    \includegraphics[scale=.4]{Objects/limitedsample_plot.pdf}
    \label{fig:limitedsample}
\end{figure}

@@

I use the general survey question about each hospital's EHR use to define my key treatment variable. The survey asks the extent to which the hospital utilizes an EHR and allows for three answers: not at all, partially utilize an EHR, and fully utilize an EHR. I create a binary variable equal to one if the answer is fully utilize, zero otherwise. The data will be aggregated up to the physician level due to physician level outcome variables, so I first transform the pair-level EHR variable into a treatment variable at the physician level. I create a variable capturing the first year that at least one of a physician's connected hospitals uses an EHR. For the endogeneity discussion in Section \ref{sec:endog}, I create a similar variable but only count a physician as exposed if a hospital categorized as "low-integration" adopts an EHR. The level of vertical integration between hospital and physician is defined as in \citeauthor{dynan1998assessing} (\citeyear{dynan1998assessing}) using the organization type of the hospital. Hospitals classified as IPA or PHO are low-integration, and any other type is high0integration. I also create a variable that captures the fraction of hospitals a physician works with that use an EHR, which varies on the dimension of EHR use and on number of hospitals worked with. Finally, the last variable I create before aggregating to the physician level is an indicator for whether the physician works with the same set of hospitals through the entire sample. I create this variable by comparing a physician's number of hospitals in each year to the maximum number of hospitals they ever work with. If there is any year in which the a physician's number of hospitals is less than the maximum number of hospitals they ever work with, the indicator for never having a new NPI is set to 0. I use this variable to limit the sample in Section \ref{sec:patientcount}.

@@

The main specification of the paper defines a binary treatment variable for whether a physician is exposed to an EHR in any hospital they work in. These results are biased if physicians endogenously select EHR exposure, particularly by participating in the hospital's decision to implement. The results do not indicate that this is a huge issue on average, since we would expect physicians not to respond on the margin of retirement or switching work setting if they are choosing to be exposed to EHRs. Especially in larger hospitals, this is unlikely to occur as an individual physician will not have much input to high level decisions. However, as an attempt to provide more evidence of labor market changes as a result of EHR exposure, I provide results for a specification with an alternate binary treatment variable: whether a physician has been exposed to an EHR in a hospital considered to have low vertical integration with physicians.

I define a hospital's level of integration based on organizational structure, where independent practice associations (IPAs) and physician-hospital organizations (PHOs) are classified as low vertical integration and other organization types are classified as high vertical integration (\cite{dynan1998assessing}).

I present a comparison of the average treatment effect in the main specification and the average treatment effect with low-integration exposure as the binary treatment variable in Figure \ref{fig:LI}. For conciseness, I present group time treatment effects to a single ATT value. As expected, the coefficient on retirement is nearly identical to the main specification. However, the coefficient for the probability of working in office based settings completely switches signs, pointing to questions about acquisition of physician practices. In low-integration settings, physicians are less likely to shift to an office after EHR exposure by a small measure. However, when including physicians with more power, the do seem to start office based work. If high-integration hospitals are acquiring smaller practices as EHRs are implemented, physicians from the hospital may be sent to the newly acquired practice for a period of time during transition. This is an important result because it points to physicians shifting according to hospital incentives instead of as a method to avoid EHR implementation. Estimates for other coefficients are more noisy, but similar to the main specification.

@@

\begin{figure}
    \centering
    \caption{Comparison of ATT, Main Specification and Low-Integration Treatment Variable}
    \includegraphics[scale=.5]{Objects/LI_results.pdf}
    \label{fig:LI}
\end{figure}
